---
title: ""scRNA-seq QC, Doublet & Contaminants Filtering"
output: html_notebook
---

### P0352 
```{r,  echo=FALSE}
set.seed(1234)
library(Seurat)
library(ggplot2)
```


```{r, echo=FALSE}
#P0352 J0 BMZ  B cells  
P0352_BMZ_Bcells <- Read10X_h5 ("C:/Users/209352/OneDrive - UPEC/SC_aCD40_BCR_vaccins/FRENCHCOVID/GEX_filtered_feature_bc_matrix.h5/0352/BMZ_0352_sample_filtered_feature_bc_matrix.h5", use.names = TRUE, unique.features = TRUE)

```
##Create and view seurat objects 

###P0352 J0 BMZ  B cells
```{r, echo=FALSE}
P0352_BMZ_Bcells <- CreateSeuratObject(counts = P0352_BMZ_Bcells, project = "P0352_BMZ", min.cells = 3, min.features = 200)
P0352_BMZ_Bcells
```



```{r, echo=FALSE}
cols0 <-  c(
    "P0352_Naive" = "#0047AB",
    "P0352_BMZ" = "#FFC300",
    "P0352_Spikepos" = "#C70039",
    "P0352_Memory" = "#008000",

    "P0352_Novavax_CTVpos" = "yellow",
    "P0352_Novavax_CTVneg" = "lightgreen",
    "P0352_Sanofi_CTVpos" = "#FF4500",   # orange rouge (Sanofi pos)
  "P0352_Sanofi_CTVneg" = "#DDA0DD",  
    "P0352_CD40cov2_CTVpos" = "purple",
    "P0352_CD40cov2_CTVneg" = "lightblue"
  )

```




### QC and selecting cells for further analysis
```{r, echo=FALSE}
P0352_BMZ_Bcells <- PercentageFeatureSet(P0352_BMZ_Bcells, pattern = "^MT-", col.name = "percent.mt")
range(P0352_BMZ_Bcells$percent.mt)
```
### Visualize QC metrics as a violin plot
```{r, echo=FALSE}

vln1 <- VlnPlot(P0352_BMZ_Bcells, features = c("nCount_RNA", "nFeature_RNA", "percent.mt"), cols= cols0, ncol = 3)
vln1
ggsave(paste0("C:/Users/209352/OneDrive - UPEC/SC_aCD40_BCR_vaccins/FRENCHCOVID/GEX_filtered_feature_bc_matrix.h5/0352/", "Figures/vln_metrics_10_04_2026.png"),vln1, width = 8, height = 5, dpi = 1000)
```
```{r, echo= FALSE}
P0352_BMZ_Bcells <- subset(P0352_BMZ_Bcells, subset = nFeature_RNA > 500 & nFeature_RNA < 3500 & nCount_RNA < 15000 & percent.mt < 5)
P0352_BMZ_Bcells
```




#### Normalisation et PCA (si ce n'est pas encore fait)
```{r, echo= FALSE}

P0352_BMZ_Bcells_QC <- NormalizeData(P0352_BMZ_Bcells)
P0352_BMZ_Bcells_QC <- FindVariableFeatures(P0352_BMZ_Bcells_QC)
P0352_BMZ_Bcells_QC <- ScaleData(P0352_BMZ_Bcells_QC)
P0352_BMZ_Bcells_QC <- RunPCA(P0352_BMZ_Bcells_QC)
```

```{r}
Elbow <- ElbowPlot(P0352_BMZ_Bcells_QC, ndims = 50) 
Elbow
ggsave(paste0("C:/Users/209352/OneDrive - UPEC/SC_aCD40_BCR_vaccins/FRENCHCOVID/GEX_filtered_feature_bc_matrix.h5/0352/", "Figures/Elbowplot_10_04_2026.png"),Elbow, width = 8, height = 5, dpi = 1000)
```

```{r, echo= FALSE}

P0352_BMZ_Bcells_QC <- FindNeighbors(P0352_BMZ_Bcells_QC, dims = 1:20)

```

```{r, echo=FALSE}

P0352_BMZ_Bcells_QC <- RunUMAP(P0352_BMZ_Bcells_QC, dims = 1:20)

```
```{r, echo=FALSE}
P0352_BMZ_Bcells_QC <- FindClusters(P0352_BMZ_Bcells_QC, resolution = 0.5)
```
```{r}
DimPlot(P0352_BMZ_Bcells_QC, reduction = "umap", group.by = "orig.ident" , cols = cols0, label = TRUE)

```

#### Find optimal parameter settings for DoubletFinder
```{r, echo = FALSE}
library(DoubletFinder)


# Use on P0352_BMZ_Bcells_QC after NormalizeData()
#                        FindVariableGenes()
#                        ScaleData()
#                        RunPCA() 
#                        RunTSNE()
# PCs: number of PCs used in Seurat to calculate clusters

# Exemple générique (à adapter selon ton objet)

sweep.res <- paramSweep(P0352_BMZ_Bcells_QC,PCs=1:20)
```


```{r, echo = FALSE}
# Summarize results from paramSweep 
# GT: Ground truth doublet controls included?
sweep.stats <-summarizeSweep(sweep.res,GT=FALSE)

```


## Select value for pK parameter
### pK defines size of cluster around the cells
### do artificial doblets belong to the cluster?
### Ignore NULL message
```{r, echo = FALSE}
bcmvn <- find.pK(sweep.stats)
head(bcmvn)
```

### find the pK with the highest BCmetric
```{r, echo = FALSE, fig.width= 10, fig.height= 6}
library(ggplot2)
ggplot(bcmvn, aes(x = pK, y = BCmetric)) +
  geom_point() + geom_line() +
  theme_minimal() +
  labs(title = "Optimization of pK parameter",
       x = "pK", y = "BCmvn metric")
```


```{r, echo = FALSE}
pKf <- bcmvn[which.max(bcmvn$BCmetric),"pK"]
pKf
pK <- as.numeric(levels(pKf))[pKf]
pK
```

### Estimate number of expected doublets
#####(assume 10% of cells are doublets)
 
```{r, echo = FALSE}

nrCells <- length(Cells(P0352_BMZ_Bcells_QC))
T5pct <- 0.1 * nrCells  # 5% de doublets articiels
# T10pct <- 0.10 * nrCells
P0352_BMZ_Bcells_QC <- doubletFinder(P0352_BMZ_Bcells_QC, pK=pK, nExp=T5pct, PCs=1:20)

```



## Summarize Results
The meta data now contain new columncontaining the results of the runs: cells that are considered doublets are labeled "doublet", the other cells are labeled "singlet":  

```{r}
head(P0352_BMZ_Bcells_QC@meta.data)
```


### Doublets found in the first run?
The results for the first run are stored in the one but last column of the meta data.  

```{r}
nc <- ncol(P0352_BMZ_Bcells_QC@meta.data)
res1 <- P0352_BMZ_Bcells_QC@meta.data[,nc]
head(P0352_BMZ_Bcells_QC@meta.data[res1 == "Doublet",])
```



##we will create a new column in the meta data called 'doublet.class'. 

```{r}
P0352_BMZ_Bcells_QC@meta.data$doublet.class <- "Singlet"
head(P0352_BMZ_Bcells_QC@meta.data$doublet.class)
P0352_BMZ_Bcells_QC@meta.data$doublet.class[res1 == "Doublet"] <- "Doublet"
table(P0352_BMZ_Bcells_QC@meta.data$doublet.class)
```


## Create plot
We will visualize the doublets on a UMAP plot:  

```{r, echo = FALSE}
umap_doublets <- DimPlot(P0352_BMZ_Bcells_QC,
        reduction="umap",
        pt.size=1.5,
        group.by="doublet.class",
        order=c("Singlet","Doublet"))
umap_doublets
ggsave(paste0("C:/Users/209352/OneDrive - UPEC/SC_aCD40_BCR_vaccins/FRENCHCOVID/GEX_filtered_feature_bc_matrix.h5/0352/", "Figures/umap_doublets_10_04_2026.png"),umap_doublets, width = 8, height = 5, dpi = 1000)


DimPlot(P0352_BMZ_Bcells_QC,
        reduction="umap",
        pt.size=1.5,
        group.by="orig.ident", cols = cols0)
```
```{r, echo=FALSE}
# Supprimer les cellules annotées comme "Doublet"
P0352_BMZ_Bcells_QC_filtered <- subset(P0352_BMZ_Bcells_QC, subset = doublet.class != "Doublet")
P0352_BMZ_Bcells_QC_filtered
```



### delete Doublet cells in the non normalized seurat object
```{r, echo=FALSE}
P0352_BMZ_Bcells_filtered <- subset(
  P0352_BMZ_Bcells,
  cells = colnames(P0352_BMZ_Bcells_QC_filtered)
)

P0352_BMZ_Bcells_filtered
```

### Marqueurs principaux (B cells)
```{r, echo=FALSE, fig.width= 10, fig.height= 7}

FeaturePlot(P0352_BMZ_Bcells_QC_filtered, features = c("CD19","MS4A1","CD79A","CD79B", "IGHM", "IGHD", "IGHG1", "IGHG2", "IGHG3", "IGHG4", "IGHA1", "IGHA2")) # B cells
FeaturePlot(P0352_BMZ_Bcells_QC_filtered, features = c("IGHM", "IGHD", "IGHG1", "IGHG2", "IGHG3", "IGHG4", "IGHA1", "IGHA2", "IGHE")) # B cells heavy chain
FeaturePlot(P0352_BMZ_Bcells_QC_filtered, features = c("IGKC","IGKC", "IGLC1","IGLC2","IGLC3","IGLC6","IGLC7")) # B cells features_light chain
```
### Visualize QC metrics as a violin plot
```{r, echo=FALSE}
####verification
vln2 <- VlnPlot(P0352_BMZ_Bcells_filtered, features = c("nCount_RNA", "nFeature_RNA", "percent.mt"), cols= cols0, ncol = 3)
vln2
ggsave(paste0("C:/Users/209352/OneDrive - UPEC/SC_aCD40_BCR_vaccins/FRENCHCOVID/GEX_filtered_feature_bc_matrix.h5/0352/", "Figures/vln_metric_after_10_04_2026.png"),vln2, width = 8, height = 5, dpi = 1000)
```

```{r}
# P0352_BMZ_Bcells_filtered <- subset(P0352_BMZ_Bcells_filtered, subset = nFeature_RNA > 500 & nFeature_RNA < 850 & nCount_RNA < 1800 & percent.mt < 6)
# P0352_BMZ_Bcells_filtered
```


```{r, echo= FALSE}
FeaturePlot(P0352_BMZ_Bcells_QC_filtered, features = c("CD3D","CD3E"))                  # T cells
FeaturePlot(P0352_BMZ_Bcells_QC_filtered, features = c("NKG7","GNLY"))                  # NK cells
FeaturePlot(P0352_BMZ_Bcells_QC_filtered, features = c("CD14","LYZ"))                   # Monocytes
#FeaturePlot(P0352_BMZ_Bcells_QC_filtered, features = c("MKI67"))                    # Prolifération
```


```{r, echo= FALSE}
library(Seurat)
library(patchwork)

p1 <- FeaturePlot(P0352_BMZ_Bcells_QC_filtered, features = c("CD3D", "CD3E"))
p2 <- FeaturePlot(P0352_BMZ_Bcells_QC_filtered, features = c("NKG7", "GNLY"))
p3 <- FeaturePlot(P0352_BMZ_Bcells_QC_filtered, features = c("CD14", "LYZ"))

combined_plot <- p1 / p2 / p3
ggsave(paste0("C:/Users/209352/OneDrive - UPEC/SC_aCD40_BCR_vaccins/FRENCHCOVID/GEX_filtered_feature_bc_matrix.h5/0352/", "Figures/combined_plot_contaminants_10_04_2026.png"),combined_plot, width = 5, height = 7, dpi = 1000
)
```


```{r, echo= FALSE}
p <- DimPlot(P0352_BMZ_Bcells_QC_filtered, reduction = "umap")
cells.to.rm <- CellSelector(p)  # dessine un polygone autour des cellules T et mono elle n'exprimment pas de BCR

```

### Supprime les cellule contaminantes 
```{r, echo= FALSE}
# Supprime les cellules sélectionnées
P0352_BMZ_Bcells_filtered <- subset(
  P0352_BMZ_Bcells_filtered,
  cells = setdiff(colnames(P0352_BMZ_Bcells_QC_filtered), cells.to.rm))

P0352_BMZ_Bcells_filtered
```

```{r, echo= FALSE}
table(P0352_BMZ_Bcells_filtered@meta.data$orig.ident)

save(P0352_BMZ_Bcells_filtered, file = "C:/Users/209352/OneDrive - UPEC/SC_aCD40_BCR_vaccins/FRENCHCOVID/GEX_filtered_feature_bc_matrix.h5/0352/Files/P0352_BMZ_Bcells_filtered.RDATA")
```




